Weakly Nonlinear MRI: Notes

### 2014-02-12 notes/mri_notes.tex
fixed lorentz force term
more
better pressure scaling explanation

And thus: $\Omega_0 r_0 \delta u = \frac{\delta}{\epsilon} \sqrt{\frac{P_0}{\rho_0}}$ \\

@@

$\partial_t u_z \, + \, \left(\mathbf{u} \cdot \nabla\right)u_z \,  =  \, - \, \frac{\epsilon^2}{\delta^2} \frac{1}{\rho} \partial_z P \, +\, \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{4 \pi \rho} \left( \nabla \times \mathbf{B} \right) \times \mathbf{B} \, + \, \frac{1}{R_e}\nabla^2 u_z $ \\

@@

@@

$\frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{4 \pi \rho} \left( \nabla \times \mathbf{B} \right) \times \mathbf{B} \, = \, \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{4 \pi \rho} \left(-\frac{\nabla B_0^2}{2} \,  - \, \frac{\mathbf{B_1}^2}{2} \, + \, \mathbf{B_1}\cdot \nabla \mathbf{B_1} \, \right)$ \\

These become: \\

$\partial_t u_{1x} \, - \, q \Omega_0 x \partial_y u_{1x} \,+ \, \left(\mathbf{u_1} \cdot \nabla\right)u_{1x} \, + \, 2 q \Omega_0 x \, - \, 2 u_{1y}\,= \, 2 q x \, - \, \frac{\epsilon^2}{\delta^2} \left( \frac{\partial_x P_1}{\rho_1 + \rho_b} - \frac{\rho_1\partial_x P_b}{\rho_b^2 + \rho_1\rho_b}\right) \, -\, \frac{1}{8\pi} \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{\rho} \partial_x \left(\mathbf{B}^2\right) \, + \, \frac{1}{4\pi} \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{\rho} \left(\mathbf{B} \cdot \nabla\right) B_x \, + \, \frac{1}{R_e}\nabla^2 u_x  $ \\

>>>>

And thus: $\Omega_0 r_0 \delta = \frac{\delta}{\epsilon} \sqrt{\frac{P_0}{\rho_0}}$ \\

Why is this a reasonable choice for the pressure scaling? The disk is vertically pressure-supported. The height of the disk is determined by the sound speed and gravity. At the radius of our box, $r_0$, the vertical scale is $h_0 = c_s/\Omega_0$. Thus $\epsilon r_0 = c_s/\Omega_0$ and $\epsilon$ is small if the Keplerian velocity is highly supersonic. This scaling allows us to substitute out the sound speed, which is meaningless in an incompressible flow anyway. \\

@@

$\partial_t u_z \, + \, \left(\mathbf{u} \cdot \nabla\right)u_z \,  =  \, - \, \frac{\epsilon^2}{\delta^2} \frac{1}{\rho} \partial_z P \,  + \, \frac{\epsilon^2}{\delta^2} \frac{1}{\rho_b} \partial_z P_b \, +\, \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{4 \pi \rho} \left( \nabla \times \mathbf{B} \right) \times \mathbf{B} \, + \, \frac{1}{R_e}\nabla^2 u_z $ \\

@@

$ = \, \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{4 \pi \rho} \left(-\frac{\nabla B_0^2}{2} \, - \, B_0 \mathbf{\hat{z}} \cdot \nabla \mathbf{B_1} \, + \, B_0 \cdot \nabla \mathbf{B_1} \,+ \, B_0 \mathbf{\hat{z}} \cdot \nabla \mathbf{B_1} \, - \, \frac{\mathbf{B_1}^2}{2} \, + \, \mathbf{B_1}\cdot \nabla \mathbf{B_1} \, \right)$ \\

@@

$\frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{4 \pi \rho} \left( \nabla \times \mathbf{B} \right) \times \mathbf{B} \, = \, \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{4 \pi \rho} \left(B_0 \mathbf{\hat{z}} \cdot \nabla \mathbf{B_1} - \, \frac{\mathbf{B_1}^2}{2} \, + \, \mathbf{B_1}\cdot \nabla \mathbf{B_1} \, \right)$ \\

Which simplifies to: \\

$\frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{4 \pi \rho} \left( \nabla \times \mathbf{B} \right) \times \mathbf{B} \, = \, \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{4 \pi \rho} \left(B_0 \partial_z \mathbf{B_1} - \, \frac{\mathbf{B_1}^2}{2} \, + \, \mathbf{B_1}\cdot \nabla \mathbf{B_1} \, \right)$ \\

The three momentum equations become: \\

$\partial_t u_{1x} \, - \, q \Omega_0 x \partial_y u_{1x} \,+ \, \left(\mathbf{u_1} \cdot \nabla\right)u_{1x} \, + \, 2 q \Omega_0 x \, - \, 2 u_{1y}\,= \, 2 q x \, - \, \frac{\epsilon^2}{\delta^2} \left( \frac{\partial_x P_1}{\rho_1 + \rho_b} - \frac{\rho_1\partial_x P_b}{\rho_b^2 + \rho_1\rho_b}\right) \, + \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{4\pi} \frac{1}{\rho_b + \rho} \left( B_0 \partial_z B_{1x} \,- \, \frac{B_{1x}^2}{2} \, + \, \mathbf{B_1}\cdot \nabla B_{1x} \right) \, + \, \frac{1}{R_e}\nabla^2 u_x  $ \\

$\partial_t u_{1y} \, - \, q \Omega_0 x \partial_y u_{1y} \,+ \, \left(\mathbf{u_1} \cdot \nabla\right)u_{1y} \,- \, 2 q \Omega_0 x \, + \, 2u_{1y} \, =  $\\
$\, - \, \frac{\epsilon^2}{\delta^2} \left( \frac{\partial_y P_1}{\rho_1 + \rho_b} \right) \, + \, \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{4\pi} \frac{1}{\rho_b + \rho} \left( B_0 \partial_z B_{1y} \,- \, \frac{B_{1y}^2}{2} \, + \, \mathbf{B_1}\cdot \nabla B_{1y} \right) \, + \, \frac{1}{R_e}\nabla^2 u_y  $ \\

$\partial_t u_{1z} \, - \, q \Omega_0 x \partial_y u_{1z} \,+ \, \left(\mathbf{u_1} \cdot \nabla\right)u_{1z} \, = $\\
$\, - \, \frac{\epsilon^2}{\delta^2} \left( \frac{\partial_z P_1}{\rho_1 + \rho_b} - \frac{\rho_1\partial_z P_b}{\rho_b^2 + \rho_1\rho_b}\right) \, + \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{4\pi} \frac{1}{\rho_b + \rho} \left( B_0 \partial_z B_{1z} \,- \, \frac{B_{1z}^2}{2} \, + \, \mathbf{B_1}\cdot \nabla B_{1z} \right) \, + \, \frac{1}{R_e}\nabla^2 u_x  $ \\

In the small shearing box (SSB) limit, $\delta \ll \epsilon \ll 1$. For a rotation law that is close to Keplerian,
